NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
edits to methods (simulation)

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python program ({\tt scripts/calcOrderParameters.py}) that uses the
MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

The order parameters of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
and can therefore be used to measure the ion binding affinity.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup order parameters at varying cation
concentrations can be easily calculated from
simulations and compared to experimental data~\cite{catte16}. The headgroup order parameters
of negatively charged PS and PG lipids also exhibit systematic, but less
understood dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, measuring the PC headgroup order parameters from 
mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see SI section \ref{electrometerFORmixtures} provides a more straightforward way of characterizing the ion binding to negatively charged membranes.
% Removed a paragraph here because it was out of place in the molecular electrometer section and contained some results. I  rewrote it and added to the end of next section

Important preliminary step for using the molecular electrometer concept is to calibrate the PC order parameter response to a known amount of bound charge~\cite{catte16,melcr18}.
This can be done using experimental data from mixtures of
monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89,melcr18},
(see SI section \ref{electrometerCALIBRATION}). Additionally, we quantify the response of PC headgroup order parameters
to the negatively charged PS, which also follows the molecular electrometer
concept in experiments~\cite{scherer87} (see SI section \ref{electrometerFORmixtures}).

Studies applying the molecular electrometer concept have utilized two different definitions for salt concentration:
The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16},
or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the reference
experimental data \cite{roux90}. However, the choice of definition has only a marginal effect
to the results in simulations with realistic ion binding affinity
(see SI section \ref{concentrationDEFsection}).

>>>>

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were calculated by first averaging over time separately
for each lipid in the system, and then calculating the average and
the standard error of the mean over the different lipids. The analysis was conducted using a
Python program ({\tt calcOrderParameters.py}, available in Ref. \citenum{MATCHgit}) that utilizes the
MDAnalysis library \cite{agrawal11,gowers16}. 
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

The order parameters of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
and can therefore be used to measure the ion binding affinity.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup order parameters at varying cation
concentrations can be easily calculated from
simulations~\cite{catte16}. The headgroup order parameters
of negatively charged PS and PG lipids also exhibit systematic, but less
understood dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, measuring the PC headgroup order parameters from 
mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.

Calibrating the PC order parameter response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer concept.
This can be done using experimental data from mixtures of
monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89,melcr18},
(see SI section \ref{electrometerCALIBRATION}). Additionally, we quantify the response of PC headgroup order parameters
to the negatively charged PS, which also follows the molecular electrometer
concept in experiments~\cite{scherer87} (see SI section \ref{electrometerFORmixtures}).

Studies applying the molecular electrometer concept have utilized two different definitions for salt concentration:
the concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16},
or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the reference
experimental data \cite{roux90}. However, the choice of definition has only a marginal effect
to the results in simulations with realistic ion binding affinity
(see SI section \ref{concentrationDEFsection}).
